---
title: "Network Analysis in R (NSCR-R Workshop)"
output: html
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries

```{r, warning = F, message = F}
library(tidyverse)
library(tidygraph)
library(igraph)
library(ggraph)
library(statnet)
```

# 1. The Basics: Network Construction and Network Visualization

__London gang network data from: __ http://www.socialsciences.manchester.ac.uk/mitchell-centre/research/covert-networks (via https://github.com/schochastics/networkdata)  

### 1.1 One-mode network from an adjacency matrix  

```{r}
# Read a matrix in which cells indicate who is connected to whom: 
m <- read.csv("https://raw.githubusercontent.com/Ieke/NSCR_network_workshop/main/matrix_example.csv", header=TRUE, row.names=1, check.names = F) %>%   
  as.matrix()
dim(m) 
```

```{r}
# Create a network object using igraph 
gang.network1 <- graph_from_adjacency_matrix(m, mode=c("undirected"), weighted=NULL)
# Inspect the network
gang.network1
```

### 1.2 One-mode network from an edgelist and attribute data 

```{r}
# Read the edgelist in which each row represents an edge from one node to another node 
e <- read.csv("https://raw.githubusercontent.com/Ieke/NSCR_network_workshop/main/edges_example.csv", header=TRUE) 
head(e)
```

```{r}
# Edgelists exclude the isolates  
# Bring in the attribute data for all nodes (including the isolates )
a <- read.csv("https://raw.githubusercontent.com/Ieke/NSCR_network_workshop/main/attributes_example.csv", header=TRUE)
head(a)
```

```{r}
# Create a network object using igraph 
gang.network2 <- graph_from_data_frame(e, directed=FALSE, vertices=a)
gang.network2
```

```{r}
# Plot the networks side by side (worry about prettier visualizations later)
# Default layout: layout_nicely (chooses best layout). 
# Connected graphs with less than 1000 vertices: default = Fruchterman-Reingold layout  
# https://schneide.blog/tag/fruchterman-reingold/
par(mfrow=(c(1,2)))
plot(gang.network1)
plot(gang.network2)
```

### 1.3 Two-mode and one-mode networks from event data  

```{r}
# Read event data  
event <- read.csv("https://raw.githubusercontent.com/Ieke/NSCR_network_workshop/main/event_example.csv", header=T)
unique(event$event) 
unique(event$person)
```

```{r}
# Transfer event data to a network format 
g <- graph_from_data_frame(event, directed = FALSE)  
# This is a bipartite graph: vertices can be divided into two classes
# No vertices of the same class are connected by an edge (e.g. every person is connected to another VIA a co-arrest event)
bipartite_mapping(g)
```

```{r}
# Indicate type of vertex (true for person)
V(g)$type <- bipartite_mapping(g)$type   
# Create one-mode network
g_one_mode <- bipartite_projection(g, which = "true", remove.type=TRUE)  
```

```{r}
# Add a few node attributes to the bipartite graph for visualization purposes 
# see also tidygraph below for adding node attributes
V(g)$'event' <- ifelse(stringr::str_detect(V(g)$'name', "[a-zA-z]"), 1,0)
V(g)$'color' <- ifelse(V(g)$'event'==1, "grey","black")
V(g)$'shape' <- ifelse(V(g)$'event'==1, "square", "circle")
V(g)$'labelcolor' <- ifelse(V(g)$'event'==1, "black", "white")
```

```{r}
# Plot the networks side by side 
# Layout kamada.kawai: based on the distance between nodes; similar nodes with small distances move in groups away from more dissimilar vertices 
set.seed(100)
par(mfrow=(c(1,2)))
plot(g, vertex.color=V(g)$'color', 
     vertex.frame.color = "white", 
     vertex.shape = V(g)$'shape', 
     vertex.label.color=V(g)$'labelcolor', 
     layout=layout.kamada.kawai) # other examples for visualizations: circle; fruchterman.reingold 
plot(g_one_mode, 
     vertex.color = "black",
     vertex.frame.color = "white", 
     vertex.label.color="white",
     layout=layout.kamada.kawai)
```

# 2. Tidy Networks 

### 2.1 Optional: Pull London gang network data from an open repository with sample datasets 

```{r, warning = F, message = F}
# https://bernhardbieri.ch/blog/ergms-and-co-offending-an-applied-social-networks-project/ ### DELETE ME
# See: https://github.com/schochastics/networkdata
#install.packages("remotes")
#remotes::install_github("schochastics/networkdata")
library(networkdata)
#data(package="networkdata") # shows all datasets available
```

```{r}
data(covert_28) 
london_gang <- covert_28
# All cells in the adjacency matrix for 'london_gang' are identical to the one associated with our sample code above (same source)
length(as_adjacency_matrix(london_gang) == as_adjacency_matrix(gang.network2)) == max(length(as_adjacency_matrix(london_gang)))
```

```{r}
# The following code was used to create the csv files above
#m <- as_adjacency_matrix(london_gang)
#write.csv(m, matrix.csv")

#e <- as_long_data_frame(london_gang)[,1:2]
#write.csv(e,"edges.csv")

#g_df <- data.frame(list(index=V(covert_28)$name, Age=V(covert_28)$Age, Arrests=V(covert_28)$Arrests, Convictions=V(covert_28)$Convictions, Prison=V(covert_28)$Prison, Birthplace=V(covert_28)$Birthplace))
#write.csv(g_df,"attributes.csv")
```

### 2.2 London gang network data from readymade csv files 

__Source: __ Data represents a co-offending network in a London-based inner-city street gang, 2005-2009, operating from a social housing estate. Data comes from anonimized police arrest and conviction data for ‘all confirmed’ members of the gang.  

__Published in: __ 
- Grund, T. and Densley, J. (2015) Ethnic Homophily and Triad Closure: Mapping Internal Gang Structure Using Exponential Random Graph Models. Journal of Contemporary Criminal Justice, Vol. 31, Issue 3, pp. 354-370 
- Grund, T. and Densley, J. (2012) Ethnic Heterogeneity in the Activity and Structure of a Black Street Gang. European Journal of Criminology, Vol. 9, Issue 3, pp. 388-406.

```{r}
london_gang <- gang.network2
```

### 2.3 Working with tidy network objects 

```{r}
# Convert the igraph network object to a tidy network object so that we can adhere to the tidy functions we're used to from the tidyverse package  
london_gang <- as_tbl_graph(london_gang)
london_gang
```

```{r}
# A tidygraph object responds to both tidyverse and igraph functions 
class(london_gang)
```

### 2.4 Node attributes 

```{r}
# Need to remove isolates: Use the following code (there are no isolates in our data, so no changes will be occur)
london_gang <- london_gang %>%
  activate(nodes) %>%
  filter(!node_is_isolated())

# Want to activate the edges instead? 
london_gang <- london_gang %>%
               activate(edges)
```

```{r}
# Activating the nodes and add attributes  
london_gang <- london_gang %>% 
  activate(nodes) %>% 
  mutate(UK_birthplace = ifelse(Birthplace==3, 1,0)) 

# Summarize node attributes
london_gang %>% 
  activate(nodes) %>% 
  as_tibble() %>% 
  summarise_at(c("Age", "Arrests", "Convictions", "Prison","UK_birthplace"), mean)
```

### 2.5 Visualizations 

```{r}
layout <- create_layout(london_gang, layout="igraph", algorithm = "kk")
london_gang %>% 
  ggraph(layout ="igraph", algorithm = "kk") + 
  geom_edge_link(colour = "grey") + 
  geom_node_point(aes(colour = as.factor(UK_birthplace))) +
  ggtitle("London Gang") + guides(color=guide_legend("UK birthplace (1 = 'Yes')"))
```

# 3. Describing the Network Structure 

### 3.1 Degree and Betweenness Centrality

__Degree centrality__ indicates a node's centrality based on its number of connections to other nodes 

__Betweenness centrality__ indicates a node's centrality based on how often it features in the shortest paths between any two other nodes. 

```{r}
# add degree centrality and betweenness centrality  
london_gang <- london_gang %>%
  activate(nodes) %>%
  mutate(degree = centrality_degree(),
         betweenness = centrality_betweenness())

# summarize node degree centralities
london_gang %>%
  activate(nodes) %>%
  as_tibble() %>%
  summarize(variable = c("Degree", "Betweenness"),
            mean = c(mean(degree), mean(betweenness)),
            max = c(max(degree), max(betweenness)),
            median = c(median(degree), median(betweenness)))

# overall centrality in the network 
centr_degree(london_gang)$centralization
centr_betw(london_gang)$centralization
```

```{r}
# Visualize nodes by their degree centrality
plot1 <- london_gang %>% 
  ggraph(layout = 'igraph', algorithm = 'kk') + 
  geom_edge_link(colour = "grey") + 
  geom_node_point(aes(size = degree, colour = degree)) + 
  ggtitle("Degree Centrality") +
  scale_color_continuous(guide = 'legend') 

# Visualize nodes by their betweenness centrality
plot2 <- london_gang %>% 
  ggraph(layout = 'igraph', algorithm = 'kk') + 
  geom_edge_link(colour = "grey") + 
  geom_node_point(aes(size = betweenness, colour = betweenness)) + 
  ggtitle("Betweenness Centrality") +
  scale_color_continuous(guide = 'legend')  

gridExtra::grid.arrange(plot1, plot2, ncol=2)
```

```{r}
# Obtain the node attributes as a dataframe
node_df <- london_gang %>%
  activate(nodes) %>%
  mutate(
    node_color = ifelse(Arrests>=median(Arrests), "red", "darkgrey")
  ) %>% as.data.frame(nodes)  
```

```{r}
# Visualize degree betweenness centrality against each other and in comparison with a node attribute
london_gang <- london_gang %>%
  activate(nodes) %>%
  mutate(degree = centrality_degree(),
         betweenness = centrality_betweenness())

ggplot(node_df, aes(x=degree, y=betweenness)) +
    labs(x = "Degree",
         y = "Betweenness") +
    geom_point(size=1.5, color=node_df$node_color) + ggtitle("Degree and betweenness")
```

### 3.2 Summarizing networks 

```{r}
describe_network <- function(graph, graph_name){
  x = tibble(
    Density = graph.density(london_gang),
    Degree = centr_degree(london_gang)$centralization,
    Betweenness = centr_betw(london_gang)$centralization,
    Closeness = centr_clo(london_gang)$centralization,
    Transitivity = transitivity(london_gang),
    Diameter = diameter(london_gang),
    Mean_distance = mean_distance(london_gang))
  x = t(x) 
  colnames(x) <- graph_name
  return(x)
}

describe_network(london_gang, "London Gang")
```

# 4. Explaining Tie Formation: Expontential Random Graph Models

### 4.1 ERGM Package in R 

```{r, warning = F, message=F}
# Load a new package that includes the functions to estimate ERGMs 
library(statnet)
```

### 4.2 Convert igraph object to a statnet object

```{r}
london_netw <- intergraph::asNetwork(london_gang)
london_netw
```

### 4.3 Estimate a null model 

```{r}
### The edge term operates as the intercept; negative coefficient = overall low probaiblity for an edge to exist between any two nodes 
model1 <- ergm(london_netw ~ edges)  
summary(model1)
```

### 4.4 Estimate a model with node attributes 

```{r}
# Probability of tie formation between any two nodes influenced by the attributes of either node (edge attributes are possible too, when available)
model2 <- ergm(london_netw ~ edges + nodecov("Age") +  nodefactor("UK_birthplace") + nodecov("Arrests"))  
summary(model2)
```

### 4.5 Estimate a model with homophily attributes 

```{r}
# Probability of tie formation between any two nodes influenced by the (dis)similarity between nodes
# NOTE: absdiff takes the absolute difference between nodes (POSITIVE coefficient = nodes that are more dissimilar have a LOWER chance to connect)
# NOTE: nodematch evaluates whether nodes share the same binary attribute (POSITIVE coefficient = nodes that are similar have a HIGHER chance to connect)
model3 <- ergm(london_netw ~ edges + nodecov("Age") + absdiff("Age") +  nodefactor("UK_birthplace") + nodematch("UK_birthplace") + nodecov("Arrests"))  
summary(model3)
```

### 4.6 Estimate a model with network attributes 

```{r}
### geometrically weighted edgewise-shared partner distribution (GWESP)
### number of triangles an edge between any two gang members closes 
model4 <- ergm(london_netw ~ edges + nodecov("Age") + absdiff("Age") +  nodefactor("UK_birthplace") + nodematch("UK_birthplace") + esp(0))
summary(model4) 
```

### 4.7 ERGM model evaluation 

```{r}
# Examine whether the estimated network aligns with the actual structure of the network 
diagn <- gof(model4 ~ model + degree + espartners + triadcensus, control = control.gof.ergm(nsim=1000)) 
plot(diagn, plotlogodds=T)
```

